Estimating duration using survival analysis

### 2024-03-25 cis-perfect-testing.tex
proof ch 5

I analyse the \numprint{4800} episodes first detected between 10 Oct 2020 and 6 Dec 2020 inclusive, with negatives bounding the start and end time of the episode.
This period is denoted as $t = 1, \dots, T$.
A negative test before the start of the episode provides a lower bound on the episode's start time; otherwise, there is little information on its length.
Similar information comes from the negative test after the end of the episode.
A negligible number of episodes have a negative test after the end of the episode.
I refer to these episodes as \emph{detected}; in \cref{perf-test:sec:prob-undetected} I formalise the conditions for an episode to be detected.
Let $\ndet$ denote the total number of detected infections.

The cohort of individuals considered is all those in the CIS who could have had a detected test.
This requires that they had at least three tests before time $T$ (a negative prior to the episode, a positive, and a negative following the episode).
Index these individuals by $i = 1, \dots, \Ncis$ where $\Ncis$ is the total number of such individuals.

An individual $i$ has a \emph{test schedule} $\sched_i$.
The test schedule $\sched_i$ is the set of times individual $i$ is tested at, from their last test prior to time 1 (if it exists) up until time $T$.
These are all the times an infection episode can be detected at, plus one prior test.
For example, for the individual in \cref{perf-test:sec:undetected}, $\sched_i = \{ 0, 7, 14, 21, 28, 56 \}$.
Each individual has exactly one test schedule, but it is possible that $\sched_i = \sched_{i'}$ for $i \neq i'$.
I assume that the test schedules are uninformative on all quantities of interest because they are part of the study design.
Therefore, I condition on them implicitly in all probabilities that follow.

@@

Any infection episode, whether detected or not, can be fully characterised by the triplet $(b_j, e_j, i(j))$ where $b_j$ and $e_j$ are the times the episode begin and end respectively, and $i(j)$ denotes the individual the episode occurred in.
The episode's duration is $d_j = e_j - b_j + 1$ (see \cref{E-inc-prev:sec:definitions}).
For all $j$, the exact values $b_j$ and $e_j$ are latent.
If $j$ is undetected then $i(j)$ is also latent and nothing is known about $b_j$ or $e_j$.
If $j$ is detected, then $i(j)$ is known; $b_j \in [l_j^{(b)}, r_j^{(b)}]$; and $e_j \in [l_j^{(e)}, r_j^{(e)}]$.
These latter two properties are what make the data doubly interval censored.

@@

To relate the possible value for $b_j$ and $e_j$, given a test schedule $\sched_{i(j)}$, to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie are strict subsets of $T \times T$).
Figure \ref{perf-test:fig:partitionSpace} shows this graphically.
Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of the parameters of the survival function, $\vec{\vec{\theta}}$, specified in \cref{perf-test:sec:parameters-priors}.

\begin{itemize}
\item
  Admissible episodes, $\alpha_i$, which have an infection and recovery time consistent with the data observed.
  This space is empty for $i \notin \set{D}$; in this case individual $i$ had $n_{ia} = 0$ admissible episodes.
  Otherwise, if $i \in \set{D}$, then individual $i$ had $n_{ia} =1$ admissible episode.
  The admissible region is the intersection of the intervals $[l_j^{(b)}, r_j^{(b)}]$ and $[l_j^{(e)}, r_j^{(e)}]$.
  This is the purple area in \cref{perf-test:fig:partitionSpace}; the boundaries in the $x$ direction of the area are $l_j^{(b)}$ and $r_j^{(b)}$ and in the $y$ direction are $l_j^{(e)}$ and $r_j^{(e)}$.
  The shape of this region accounts for the doubly interval censored nature of the data.
  Conditional on the episode occurring in individual $i$, the episode is admissible with probability $p_{ia} = \prob((B_j, E_j) \in \alpha_i \mid \vec{\theta}, i(j) = i)$.
  \label{perf-test:def:admissible}
\item
  Undetected episodes, $\Omega_i$, which have an infection and recovery time such that they would not have tested positive.
  An unknown number, $n_{iu}$ of these occur in individual $i$.
  The undetected region is the red region in \cref{perf-test:fig:partitionSpace}, which are when the individual is infected then recovers before their next test, plus when the individual is infected before their enrolment.
  The shape of this region accounts for the pattern of possible undetected infections.
  Conditional on the episode occurring in individual $i$, the episode is undetected with probability $p_{iu} = \prob((B_j, E_j) \in \Omega_i \mid \vec{\theta}, i(j) = i)$.
\item
  Inadmissible episodes, $\beta_i$, which would be detected (if they occur) but were not.
  Therefore, we know $n_{i\bar{a}} = 0$ of these occurred in individual $i$.
  These are all remaining episodes not in the previous sets.
  Conditional on the episode occurring in these individuals, the episode inadmissible with probability $p_{i\bar{a}} = \prob((B_j, E_j) \in \beta_i \mid \vec{\theta}, i(j) = i)$.
\end{itemize}

@@

Each episode occurs in one of the classes independently and with the same probability.
Therefore, conditional on $\ntot$, the number of episodes in each class is multinomial.
Specifically:
\begin{align}
  \begin{bmatrix}
    n_{1a} \\ n_{1u} \\ n_{1\bar{a}} \\ \vdots \\ n_{\Ncis,a} \\ n_{\Ncis,u} \\ n_{\Ncis,\bar{a}}
  \end{bmatrix}
  \mid \ntot
  =
  \MN\left(
    \ntot,
    \frac{1}{\Ncis}
    \begin{bmatrix}
      p_{1a} \\ p_{1u} \\ p_{1\bar{a}} \\ \vdots \\ p_{\Ncis,a} \\ p_{\Ncis,u} \\ p_{\Ncis,\bar{a}}
    \end{bmatrix}
  \right)
  \label{perf-test:eq:multinomial}
\end{align}

In forming the region $\alpha_i$, I use the test outcomes.
This is not strictly correct because the cell probabilities in the likelihood should not depend on the data.
However, an alternative considering each combination of infection and recovery times as its own cell is possible.
The new cells would align with the inadmissible region.
Since each of these cells have an observed count of zero, they can be combined to give the same likelihood.
That is, the approach is mathematically equivalent but greatly complicates the formulae and notation.
Therefore, I present this version as it is clearer.

The undetected classes cannot be distinguished.
Therefore, I combine these classes into a single class.
The combined class has a count $\nnodet = \sum_{i=1}^{\Ncis} n_{iu}$ and probability $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} p_{iu}$.

@@

Substituting \cref{perf-test:eq:augmented-likelihood} into \cref{perf-test:eq:posterior1}:
\begin{align}
&\propto p(\vec{\theta}) \sum_{\ntot=\ndet}^\infty \left( p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i \in \set{D}} p_{ia} \right) \\
&= p(\vec{\theta}) \left( \prod_{i \in \set{D}} p_{ia} \right) \left( \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right) \label{perf-test:eq:posterior2}.
\end{align}
For convenience, define the summation term as:
\begin{align}
  \eta =  \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}.
  \label{perf-test:eq:eta}
\end{align}
The rest of this section derives expressions for each of $p_{ia}$, $p_{u}$ and $\eta$.

@@

Putting a negative binomial prior on $\ntot$ is equivalent to the following gamma-Poisson composite; its use simplifies the derivation.
\begin{align}
\ntot \mid \lambda &\dist \Poi(\lambda) \\
\lambda &\dist \GamDist(a, b)
\end{align}
where $b = r / \mu$ and $a = r$.
Hence:
\begin{align}
\eta
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot \mid \lambda) p(\lambda) d\lambda &\text{$\lambda$ explicit}\\
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \frac{\lambda^{\ntot} e^{-\lambda}}{\ntot!} p(\lambda) d\lambda &\ntot \dist \Poi\\
%&= \int \sum_{\ntot=\ndet}^\infty \frac{1}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \lambda^{\ntot-\ndet} \lambda^{\ndet} e^{-\lambda} p(\lambda) d\lambda \\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) \sum_{\nnodet=0}^\infty \frac{(\pnodet \lambda)^{\nnodet}}{\nnodet!} d\lambda &\nnodet = \ntot-\ndet\\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) e^{\lambda \pnodet} d\lambda &\text{Maclaurin series of $e$} \\
&= \int \lambda^{\ndet} e^{-\lambda(1 - \pnodet)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda &\text{Gamma pdf}\\
&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+\ndet-1} e^{-(b+1-\pnodet)\lambda} \lambda d\lambda \\
&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+\ndet)}{(b+1-\pnodet)^{a+\ndet}} &\text{Gamma pdf}\\
&\propto (b+1-\pnodet)^{-(a+\ndet)} \\
&= (r/\mu + 1 - \pnodet)^{-(r+\ndet)} &\text{sub in $\mu$ and $r$}\\
&\propto(r + \mu (1- \pnodet))^{-(r+\ndet)}.
\end{align}

@@

Next, I derive $p_{ia}$.
From \cref{perf-test:eq:full-posterior}, this is only needed for $i \in \set{D}$, that is for individuals with detected episodes.
For such an individual, let $j$ denote their detected episode.

By definition (see \cpageref{perf-test:def:admissible}), we have, for any $i \in \set{D}$:
\begin{align}
p_{ia} =& \prob((B_j, E_j) \in \alpha_i \mid \vec{\theta}, i(j) = i) \\
\alpha_i =& \{ (b, e) : l_{j}^{(b)} \leq b \leq r_{j}^{(b)} \wedge l_{j}^{(e)} \leq e \leq r_{j}^{(e)}\}.
\intertext{Omitting the conditioning on $\vec{\theta}$ and $i(j) = i$, this gives:}
p_{ia}
=& \prob \left( l_{j}^{(b)} \leq B_{j} \leq r_{j}^{(b)}, l_{j}^{(e)} \leq E_{j} \leq r_{j}^{(e)} \right) \\
=& \prob \left( l_{j}^{(e)} \leq E_{j} \leq r_{j}^{(e)} \mid l_{j}^{(b)} \leq B_{j} \leq r_{j}^{(b)} \right) \\
   &\times\prob \left( l_{j}^{(b)} \leq B_{j} \leq r_{j}^{(b)} \right) \\
=& \sum_{b = l_{j}^{(l)}}^{r_{j}^{(b)}} \prob \left( l_{j}^{(e)} \leq E_{j} \leq r_{j}^{(e)} \mid B_{j} = b \right) \prob \left(B_{j} = b \right) \\
=& \sum_{b = l_{j}^{(b)}}^{r_{j}^{(b)}} \prob \left( l_{j}^{(e)} - b + 1 \leq D_{j} \leq r_{j}^{(e)} - b + 1 \right) \prob \left(B_{j} = b \right) &\text{by def of $D_{j}$} \\
=& \sum_{b = l_{j}^{(b)}}^{r_{j}^{(b)}} \left( S_{\vec{\theta}}(l_{j}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j}^{(e)} - b + 2) \right) \prob \left(B_{j} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
\propto& \sum_{b = l_{j}^{(b)}}^{r_{j}^{(b)}} \left( S_{\vec{\theta}}(l_{j}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j}^{(e)} - b + 2) \right)
\end{align}
under the assumption of uniform probability of infection time.
This is the standard form of the likelihood for doubly interval censored data without truncation~\autocite[e.g.][]{sunEmpirical}.

@@

An episode $j$ in individual $i(j)$ is detected if and only if all the following conditions are met.
\begin{enumerate}
    \item $\exists t \in \sched_{i(j)}$ such that $b_j \leq t \leq e_j$; this condition is equivalent to having at least one positive test for the episode.
    \item $\exists t \in \sched_{i(j)}$ such that $t < b_j$.
      For individuals enrolled during the period considered ($\min \sched_{i(j)} > 0$), this ensures that the beginning of the episode is lower bounded; thereby adjusting for those who enrolled during the period being less likely to have a detected infection.
      For individuals enrolled prior to the period considered ($\min \sched_{i(j)} \leq 0$), this means that the episode was not detected prior to time 1.
    \item $\exists t \in \sched_{i(j)}$ such that $t \geq b_j$.
      This means that the test is detected before $T$.
    \item $\exists t \in \sched_{i(j)}$ such that $t > e_j$, upper bounding the end of the episode.
      For episodes detected in the period I consider, this happens a negligible number of times, only in individuals lost to follow-up; therefore, I neglect this possibility.
      For a new context, including recent infections, this condition could be relaxed by considering episodes that do not meet this criterion as right censored.
\end{enumerate}

% An episode is undetected if and only if no tests are performed during the episode or if there was no negative test prior to the episode.
% Equivalently, that the first test at or after $b$ is after $e$, or that there is no negative test prior to $b$.
To formalise this, first define $\tau_{\sched_i}(t)$ as the time until the next test at or after time $t$ in the schedule $\sched_i$:
\begin{align}
\tau_{\sched_i}(t) &= \min \{ t' \in \sched_i : t' \geq t \} - t
\end{align}
defining $\min \emptyset = \infty$; \ie $\tau_{\sched_i}(t) = \infty$ if there is no $t' \in \sched_i$ such that $t' \geq t$.
The condition that $\exists t \in \sched_{i(j)}$ such that $b_j \leq t \leq e_j$ can now be expressed as $e_j \geq b_j + \tau_{\sched_{i(j)}}(b_j)$.
Equivalently, $d_j \geq \tau_{\sched_{i(j)}}(b_j) + 1$.

@@

When constructing the prior, two aspects need consideration.
Firstly, the model structure from \cref{E-ATACCC} leads to positive correlation in the posterior estimates of the hazard, which should be propagated into this analysis.
That is, the prior used in this analysis cannot be independent across the hazards.
Secondly, the uncertainty in the estimates from \cref{E-ATACCC} is underestimated because it extrapolates beyond the data using strong model assumptions (see \cref{E-ATACCC:sec:discussion}).
%Furthermore, there are differences in the study design and laboratories used between the two studies (see \cref{E-intro:sec:studies}) which may mean that results do not generalise between the studies.
I form the prior to include this prior information into the CIS analysis two steps.

I first approximate the \cref{E-ATACCC} posterior estimate of the hazard as $\logit{\vec{h}} \dist \MNorm(\vec{\mu}_A, \matr{\Sigma}_A)$ where $\vec{h}$ is the posterior distribution of the hazards, and $\vec{\mu}_A$ and $\matr{\Sigma}_A$ are the mean and covariance matrix computed from the samples of $\logit{\vec{h}}$ derived using the method of \cref{ATACCC:sec:duration}.
Using a multivariate normal, as opposed to multiple univariate distribution for each $h_t$, preserves the correlation between the hazards.
The logit transformation maps from the $[0, 1]$ interval to the full real line.
The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC-hazard} and \cref{perf-test:fig:approximate-ATACCC-survival}).
\begin{figure}
  \thisfloatpagestyle{empty}
  % \vspace{-1cm}
  \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{cis-perfect-testing/ataccc-approximation-hazard}}
  \caption[Approximating the ATACCC posterior hazard as a logit-normal]{Comparison of the posterior estimate of the hazard from \cref{E-ATACCC} and its approximation with a logit-normal distribution (kernel density estimate smoothed) for hazards between day 4 and 15. The density for smaller hazards is concentrated very close to 0. \label{perf-test:fig:approximate-ATACCC-hazard}}
\end{figure}
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/ataccc-approximation-survival}
  \caption[Approximating the ATACCC posterior survival as a logit-normal]{Comparison of the posterior estimate of the survival from \cref{E-ATACCC} and its approximation with a logit-normal distribution (median and central 95\% interval). The survival is a function of all the hazards up until that point, therefore the correlation between the hazards must be well-approximated to have these agree. \label{perf-test:fig:approximate-ATACCC-survival}}
\end{figure}

>>>>

I analyse the $\ndet = \numprint{4800}$ episodes with their first positive test between 10 Oct 2020 and 6 Dec 2020 inclusive, with negatives bounding the start and end time of the episode; I refer to these episodes as \emph{detected}.
In \cref{perf-test:sec:prob-undetected} I formalise the conditions for an episode to be detected.
This period is denoted as $t = 1, \dots, T$.
A negative test before the start of the episode provides a lower bound on the episode's start time; otherwise, there is little information on its length.
Similar information comes from the negative test after the end of the episode.
A negligible number of episodes are missing a negative test after the end of the episode.

I consider the cohort of $\Ncis$ individuals who could have had a detected episode; this is a subset of the whole CIS cohort.
To be considered, an individual must have had at least three tests before time $T$ (a negative prior to the episode, a positive, and a negative following the episode).
I index these individuals by $i = 1, \dots, \Ncis$.

An individual $i$ has a \emph{test schedule} $\sched_i$.
The test schedule $\sched_i$ is the set of times individual $i$ is tested at, from their last test prior to time 1 (if it exists) up until time $T$.
These are all the times an infection episode can be detected at, plus one prior test.
For example, for the individual in \cref{perf-test:fig:double-interval-censor}, $\sched_i = \{ 0, 7, 14, 21, 28, 56 \}$.
Each individual has exactly one test schedule, but it is possible that $\sched_i = \sched_{i'}$ for $i \neq i'$.
I assume that the test schedules are uninformative on all quantities of interest because they are part of the study design.
Therefore, I condition on them implicitly in all probabilities that follow.

@@

Any infection episode, whether detected or not, is fully characterised for my purposes by the triplet $(b_j, e_j, i(j))$ where $b_j$ and $e_j$ are the times the episode begin and end respectively, and $i(j)$ denotes the individual the episode occurred in.
The episode's duration is $d_j = e_j - b_j + 1$ (see \cref{E-inc-prev:sec:definitions}).
For all $j$, the exact values $b_j$ and $e_j$ are latent.
If $j$ is undetected then $i(j)$ is also latent and nothing is known about $b_j$ or $e_j$.
If $j$ is detected, then $i(j)$ is known; $b_j \in [l_j^{(b)}, r_j^{(b)}]$; and $e_j \in [l_j^{(e)}, r_j^{(e)}]$.
These latter two properties are what make the data doubly interval censored.

@@

To relate the possible value for $b_j$ and $e_j$, given a test schedule $\sched_{i(j)}$, to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie are strict subsets of $T \times T$), shown in \cref{perf-test:fig:partitionSpace}.
Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of the parameters of the survival function, $\vec{\vec{\theta}}$, specified in \cref{perf-test:sec:parameters-priors}.

\begin{itemize}
\item
  Admissible episodes, $\alpha_i$, which have an infection and recovery time consistent with the data observed.
  This space is empty for $i \notin \set{D}$; in this case individual $i$ had $n_{ia} = 0$ admissible episodes.
  Otherwise, if $i \in \set{D}$, then individual $i$ had $n_{ia} =1$ admissible episode.
  The admissible region is the intersection of the intervals $[l_j^{(b)}, r_j^{(b)}]$ and $[l_j^{(e)}, r_j^{(e)}]$.
  This is the purple area in \cref{perf-test:fig:partitionSpace}; the boundaries in the $x$ direction of the area are $l_j^{(b)}$ and $r_j^{(b)}$ and in the $y$ direction are $l_j^{(e)}$ and $r_j^{(e)}$.
  The shape of this region accounts for the doubly interval censored nature of the data.
  Conditional on the episode occurring in individual $i$, the episode is admissible with probability $p_{ia} = \prob((B_j, E_j) \in \alpha_i \mid \vec{\theta}, i(j) = i)$.
  \label{perf-test:def:admissible}
\item
  Undetected episodes, $\Omega_i$, which have an infection and recovery time such that they would not have tested positive.
  An unknown number, $n_{iu}$ of these occur in individual $i$.
  The undetected region is the red region in \cref{perf-test:fig:partitionSpace}, indicating combinations of episode beginning and end times which do not span a testing time plus episodes beginning before the individual's enrolment.
  The shape of this region accounts for the pattern of possible undetected infections.
  Conditional on the episode occurring in individual $i$, the episode is undetected with probability $p_{iu} = \prob((B_j, E_j) \in \Omega_i \mid \vec{\theta}, i(j) = i)$.
\item
  Inadmissible episodes, $\beta_i$, which would be detected (if they occur) but were not.
  Therefore, we know $n_{i\bar{a}} = 0$ of these occurred in individual $i$.
  These are all remaining episodes not in the previous sets.
  Conditional on the episode occurring in these individuals, the episode inadmissible with probability $p_{i\bar{a}} = \prob((B_j, E_j) \in \beta_i \mid \vec{\theta}, i(j) = i)$.
\end{itemize}

@@

Each episode occurs in one of the classes independently and with the same probability.
Therefore:
\begin{align}
  \begin{bmatrix}
    n_{1a} \\ n_{1u} \\ n_{1\bar{a}} \\ \vdots \\ n_{\Ncis,a} \\ n_{\Ncis,u} \\ n_{\Ncis,\bar{a}}
  \end{bmatrix}
  \mid \ntot
  =
  \MN\left(
    \ntot,
    \frac{1}{\Ncis}
    \begin{bmatrix}
      p_{1a} \\ p_{1u} \\ p_{1\bar{a}} \\ \vdots \\ p_{\Ncis,a} \\ p_{\Ncis,u} \\ p_{\Ncis,\bar{a}}
    \end{bmatrix}
  \right)
  \label{perf-test:eq:multinomial}
\end{align}

In forming the region $\alpha_i$, I use the test outcomes.
This is not strictly correct because the cell probabilities in the likelihood should not depend on the data.
However, an alternative considering each combination of infection and recovery times as its own cell is possible.
The new cells would align with the inadmissible region.
Since each of these cells have an observed count of zero, they can be combined to give the same likelihood.
That is, the approach is mathematically equivalent but greatly complicates the formulae and notation.
Therefore, I present this version for clarity.

The undetected classes cannot be distinguished.
Therefore, I combine these classes into a single class.
The combined class has a count $\nnodet = \sum_{i=1}^{\Ncis} n_{iu}$ and probability parameter $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} p_{iu}$.

@@

Substituting \cref{perf-test:eq:augmented-likelihood} into \cref{perf-test:eq:posterior1}:
\begin{align}
p(\vec{\theta} \mid \na)
&\propto p(\vec{\theta}) \sum_{\ntot=\ndet}^\infty \left( p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i \in \set{D}} p_{ia} \right) \\
&= p(\vec{\theta}) \left( \prod_{i \in \set{D}} p_{ia} \right) \left( \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right) \label{perf-test:eq:posterior2}.
\end{align}
For convenience, define the summation term as:
\begin{align}
  \eta =  \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}.
  \label{perf-test:eq:eta}
\end{align}
The rest of this section derives expressions for each of $p_{ia}$, $p_{u}$ and $\eta$.

@@

Putting a negative binomial prior on $\ntot$ is equivalent to the following gamma-Poisson composite; its use simplifies the derivation.
\begin{align}
\ntot \mid \lambda &\dist \Poi(\lambda) \\
\lambda &\dist \GamDist(a, b)
\end{align}
where $b = r / \mu$ and $a = r$.
Hence:
\begin{align}
\eta
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot \mid \lambda) p(\lambda) d\lambda &\text{$\lambda$ explicit}\\
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \frac{\lambda^{\ntot} e^{-\lambda}}{\ntot!} p(\lambda) d\lambda &\ntot \dist \Poi\\
%&= \int \sum_{\ntot=\ndet}^\infty \frac{1}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \lambda^{\ntot-\ndet} \lambda^{\ndet} e^{-\lambda} p(\lambda) d\lambda \\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) \sum_{\nnodet=0}^\infty \frac{(\pnodet \lambda)^{\nnodet}}{\nnodet!} d\lambda &\nnodet = \ntot-\ndet\\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) e^{\lambda \pnodet} d\lambda &\text{Maclaurin series of $e$} \\
&= \int \lambda^{\ndet} e^{-\lambda(1 - \pnodet)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda &\lambda \dist \GamDist\\
&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+\ndet-1} e^{-(b+1-\pnodet)\lambda} \lambda d\lambda \\
&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+\ndet)}{(b+1-\pnodet)^{a+\ndet}} &\text{Gamma pdf}\\
&\propto (b+1-\pnodet)^{-(a+\ndet)} \\
&= (r/\mu + 1 - \pnodet)^{-(r+\ndet)} &\text{sub in $\mu$ and $r$}\\
&\propto(r + \mu (1- \pnodet))^{-(r+\ndet)}.
\end{align}

@@

Next, I derive $p_{ia}$.
From \cref{perf-test:eq:full-posterior}, this is only needed for $i \in \set{D}$, that is for individuals with detected episodes.
For such an individual, let $j(i)$ denote their detected episode.

By definition (see \cpageref{perf-test:def:admissible}), we have, for any $i \in \set{D}$:
\begin{align}
p_{ia} =& \prob((B_j, E_j) \in \alpha_i \mid \vec{\theta}, i(j) = i) \\
\alpha_i =& \{ (b, e) : l_{j(i)}^{(b)} \leq b \leq r_{j(i)}^{(b)} \wedge l_{j(i)}^{(e)} \leq e \leq r_{j(i)}^{(e)}\}.
\intertext{Omitting the conditioning on $\vec{\theta}$ and $i(j) = i$, this gives:}
p_{ia}
=& \prob \left( l_{j(i)}^{(b)} \leq B_{j} \leq r_{j(i)}^{(b)}, l_{j(i)}^{(e)} \leq E_{j} \leq r_{j(i)}^{(e)} \right) \\
=& \prob \left( l_{j(i)}^{(e)} \leq E_{j} \leq r_{j(i)}^{(e)} \mid l_{j(i)}^{(b)} \leq B_{j} \leq r_{j(i)}^{(b)} \right) \\
   &\times\prob \left( l_{j(i)}^{(b)} \leq B_{j} \leq r_{j(i)}^{(b)} \right) \\
=& \sum_{b = l_{j(i)}^{(l)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} \leq E_{j} \leq r_{j(i)}^{(e)} \mid B_{j} = b \right) \prob \left(B_{j} = b \right) \\
=& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} - b + 1 \leq D_{j} \leq r_{j(i)}^{(e)} - b + 1 \right) \prob \left(B_{j} = b \right) &\text{by def of $D_{j}$} \\
=& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 2) \right) \prob \left(B_{j} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
\propto& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 2) \right)
\end{align}
under the assumption of uniform probability of infection time.
This is the standard form of the likelihood for doubly interval censored data without truncation~\autocite[e.g.][]{sunEmpirical}.

@@

An episode $j$ in individual $i(j)$ is detected if and only if all the following conditions are met.
\begin{enumerate}
    \item $\exists t \in \sched_{i(j)}$ such that $b_j \leq t \leq e_j$; this condition is equivalent to having at least one positive test for the episode.
    \item $\exists t \in \sched_{i(j)}$ such that $t < b_j$.
      For individuals enrolled during the period considered ($\min \sched_{i(j)} > 0$), this ensures that the beginning of the episode is lower bounded; thereby adjusting for those who enrolled during the period being less likely to have a detected infection.
      For individuals enrolled prior to the period considered ($\min \sched_{i(j)} \leq 0$), this means that the episode was not detected prior to time 1.
    \item $\exists t \in \sched_{i(j)}$ such that $t \geq b_j$.
      This means that the test is detected before $T$.
    \item Individual $i(j)$ is tested at a time $t > e_j$, upper bounding the end of the episode.
      Note that this could be after $T$, and hence it is not necessarily the case that $t \in \sched_{i(j)}$.
      For episodes detected in the period I consider, this happens a negligible number of times, only in individuals lost to follow-up; therefore, I neglect this possibility.
      % For a new context, including recent infections, this condition could be relaxed by considering episodes that do not meet this criterion as right censored.
\end{enumerate}

% An episode is undetected if and only if no tests are performed during the episode or if there was no negative test prior to the episode.
% Equivalently, that the first test at or after $b$ is after $e$, or that there is no negative test prior to $b$.
To formalise this, first define $\tau_{\sched_i}(t)$ as the time until the next test at or after time $t$ in the schedule $\sched_i$:
\begin{align}
\tau_{\sched_i}(t) &= \min \{ t' \in \sched_i : t' \geq t \} - t
\end{align}
defining $\min \emptyset = \infty$; \ie $\tau_{\sched_i}(t) = \infty$ if there is no $t' \in \sched_i$ such that $t' \geq t$.
The first condition can now be expressed as $e_j \geq b_j + \tau_{\sched_{i(j)}}(b_j)$.
Equivalently, $d_j \geq \tau_{\sched_{i(j)}}(b_j) + 1$.

@@

When constructing the prior, two aspects need consideration.
Firstly, the model structure from \cref{E-ATACCC} leads to positive correlation in the posterior estimates of the hazard, which should be propagated into this analysis.
That is, the prior used in this analysis should have correlation between $\lambda_t$ and $\lambda_{t'}$.
Secondly, the uncertainty in the estimates from \cref{E-ATACCC} is underestimated because it extrapolates beyond the data using strong model assumptions (see \cref{E-ATACCC:sec:discussion}).
%Furthermore, there are differences in the study design and laboratories used between the two studies (see \cref{E-intro:sec:studies}) which may mean that results do not generalise between the studies.
I form the prior to include this prior information into the CIS analysis two steps.

I first approximate the \cref{E-ATACCC} posterior estimate of the hazard as $\logit{\vec{h}} \dist \MNorm(\vec{\mu}_A, \matr{\Sigma}_A)$ where $\vec{h}$ is the hazard, and $\vec{\mu}_A$ and $\matr{\Sigma}_A$ are the mean and covariance matrix computed from the samples of $\logit{\vec{h}}$ derived using the method of \cref{ATACCC:sec:duration}.
Using a multivariate normal, as opposed to multiple univariate distribution for each $h_t$, preserves the correlation between the hazards.
% The logit transformation maps from the $[0, 1]$ interval to the full real line.
The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC-hazard} and \cref{perf-test:fig:approximate-ATACCC-survival}).
\begin{figure}
  \thisfloatpagestyle{empty}
  % \vspace{-1cm}
  \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{cis-perfect-testing/ataccc-approximation-hazard}}
  \caption[Approximating the ATACCC posterior hazard as a logit-normal]{Comparison of the posterior estimate of the hazard from \cref{E-ATACCC} and its approximation with a logit-normal distribution (kernel density estimate smoothed) for hazards between day 4 and 15. The density for smaller hazards is concentrated very close to 0. \label{perf-test:fig:approximate-ATACCC-hazard}}
\end{figure}
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/ataccc-approximation-survival}
  \caption[Approximating the ATACCC posterior survival as a logit-normal]{Comparison of the posterior estimate of the survival from \cref{E-ATACCC} and its approximation with a logit-normal distribution (median and central 95\% interval). The survival is a function of all the hazards up until that point, therefore the correlation between the hazards must be well-approximated to have these agree. \label{perf-test:fig:approximate-ATACCC-survival}}
\end{figure}

Having approximated the estimate as a multivariate normal, I add additional uncertainty using a discrete Beta process.
The discrete Beta process prior~\autocite{ibrahimBayesian,sunStatisticala} generalises the form of prior used in \cref{perf-test:sec:independent-priors} by allowing the central estimate of the hazard to vary over time.
It is:
\begin{align}
  \logit \vec{h} &\dist \MNorm(\vec{\mu}_A, \matr{\Sigma}_A) \\
  \lambda_t &\dist \text{Beta}(\alpha_t, \beta_t) &t = 1, 2, \dots \\
  \alpha_t &= k_t h_t + \alpha_0 \\
  \beta_t &= k_t (1 - h_t) + \beta_0
\end{align}
where $k_t$, $\alpha_0$, and $\beta_0$ are hyperparameters; and, as before, $h_t$ is an estimate of the hazard at time $t$ from \cref{E-ATACCC}.
An intuition for what this distribution represents derives from a conjugate model for $\lambda_t$ with a beta prior and a binomial likelihood.
If $\lambda_t$ is given the prior distribution $\text{Beta}(\alpha_0, \beta_0)$, and we then have $k_t$ observations with $k_t h_t$ successes, then the posterior distribution for $\lambda_t$ is $\text{Beta}(\alpha_t, \beta_t)$ (as defined above).
